Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-10 ltx/data_paper/DataPaper.tex
edits to EM section

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=\mathbf E_{\boldsymbol\Theta'}\left[\sum_{t=0}^{T-1}2\ln p(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon})+\sum_{t=0}^{T-1}2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta,\xi, \sigma_d)+\sum_{t=0}^{T-1}2\ln p(\mathbf x_0)\right].

@@

 It should be noted that the last term is not a function of parameter set and can be dropped, therefore $\mathcal Q$-function can be rewritten as

@@

Gaussian distributions of the state disturbance and the observation noise results into the following conditional distributions, where the normalising factors are ignored 

@@

p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}} \exp \left(-\frac{1}{2}(\mathbf x_{t+1}-q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t) \right)

@@

Note that the disturbance covariance matrix can be written as a product of the unknown parameter, $\sigma_d^2$ and a constant matrix, $\tilde{\boldsymbol\Sigma}_e$ where the constant part is calculated using the estimated support from the correlation analysis. By expanding the exponent and taking $2\ln$ we get

@@

2\ln p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&=-n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})\right],  \\
2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&=-n_x\ln\sigma_d^2-\ln\mid\tilde{\boldsymbol\Sigma}_e\mid-\frac{1}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_{t+1}+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\boldsymbol\theta+\frac{2\xi}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t \nonumber \\
&-\frac{1}{\sigma_e^2}\boldsymbol\theta^\top q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\frac{\xi^2}{\sigma_d^2}\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t. \label{eq:CondititionDist2}

@@

\dean{Dean to check expansion.} Now we rearrange the equation so we can  use the information from the smoother. (or isolate (reason for doing this step is..)) the cross-covariance etc. Taking trace and rearranging, using the invariant cyclic permutations property of the trace, equations \eqref{eq:CondititionDist1} and \eqref{eq:CondititionDist2} become

@@

  2\ln p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right) &= -n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\right\rbrace    \\
 2 \ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d) &=-n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace\mathbf x_{t+1} \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\boldsymbol\theta+\frac{2\xi}{\sigma_d^2} \mathrm{tr} \left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\frac{\xi^2}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf x_t \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace. \label{eq:Qfunctiontrace2}

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&= -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right]\boldsymbol\theta

@@

&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \boldsymbol\theta

@@

The second term in \eqref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation of the second term in \eqref{eq:Qfunctiontrace1}. By defining $\Xi$-variables as 

@@

\boldsymbol\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:Xi1} \\
\boldsymbol\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\label{eq:Xi2}\\
\boldsymbol\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right] \label{eq:Xi3} \\	
\boldsymbol\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right] \label{eq:Xi4}\\
 \boldsymbol\Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \label{eq:Xi5}

@@

equation \eqref{eq:Voldermont} can be written as

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \boldsymbol\Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\boldsymbol\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 

@@

&+\frac{2}{\sigma_d^2}\boldsymbol\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \boldsymbol\Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2}\boldsymbol \Xi_{5} \boldsymbol\theta.

@@

Covariance and cross-covariances in equation \eqref{eq:VoldermontwitXivariables}, consist of both linear terms, $\boldsymbol\Xi_0$, $\boldsymbol\Xi_1$ and $\boldsymbol\Xi_2$ and non-linear terms, $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and  $\boldsymbol\Xi_5$. Finding linear $\boldsymbol\Xi$-variables is straightforward as they can be computed directly using smoother outputs, i.e. $\hat{\mathbf x}_t^b $, $\mathbf P_t^b$ and $\mathbf M_t^b$ using following expressions \cite{Gibsona2005} (see supporting information for details)

@@

\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:Xi1Gibson}, \\

@@

However calculation of non-linear $\boldsymbol\Xi$-variables is not straightforward as in linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in  calculations of $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and $\boldsymbol\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^b$ at each time instant, i.e.

@@

 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^b\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)

@@

\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)\right)

@@

substituting \eqref{eq:SigmoidTaylor} in \eqref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the smoothed state $\hat{\mathbf{x}}_t^b$ 
 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t^b\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right) d\mathbf{r}'

>>>>

\begin{align}\label{eq:EM-QwithInitialstates}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=\mathbf E_{\boldsymbol\Theta'}\left[\sum_{t=0}^{T-1}2\ln p(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon})+\sum_{t=0}^{T-1}2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta,\xi, \sigma_d) \right.\nonumber \\
&\left.+\sum_{t=0}^{T-1}2\ln p(\mathbf x_0)\right].

@@

 The last term on the right-hand side of \eqref{eq:EM-QwithInitialstates} can be omitted as it does not depend on the parameter set, therefore $\mathcal Q$-function can be rewritten as

@@

Gaussian distributions of the state disturbance and the observation noise results into the following conditional distributions, where the normalising factors are ignored since scaling by a constant coefficient does not alter the location of the maximum with respect to the parameters.

@@

p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}} \nonumber \\
&\times\exp \left(-\frac{1}{2}(\mathbf x_{t+1}-\mathbf q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-\mathbf q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t) \right).

@@

Note that disturbance covariance matrix, $\boldsymbol\Sigma_e$, can be written as the product of the unknown parameter, $\sigma_d^2$ with the constant matrix, $\tilde{\boldsymbol\Sigma}_e$, where the constant part can be  calculated using the estimated support for disturbance covariance function, $\gamma$ from the correlation analysis (see equation \eqref{eq:EM-MMGFourier2}). By expanding the exponent and taking $2\ln$ we get

@@

2\ln p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&=-n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})\right],

@@

2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&=-n_x\ln\sigma_d^2-\ln\mid\tilde{\boldsymbol\Sigma}_e\mid-\frac{1}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_{t+1}\nonumber \\
&+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta 
+\frac{2\xi}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t \nonumber \\
&-\frac{1}{\sigma_e^2}\boldsymbol\theta^\top \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta\nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t. \label{eq:CondititionDist2}
In the next step the above equation is rearranged in order to isolate the covariance and the cross-covariance terms, enabling the use of smoother outputs. Taking trace and rearranging, applying the invariant cyclic permutations property of the trace, equations \eqref{eq:CondititionDist1} and \eqref{eq:CondititionDist2} become

@@

  2\ln p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right) &= -n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\right\rbrace    
 2 \ln &p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d) =-n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace\mathbf x_{t+1} \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta \nonumber \\
&+\frac{2\xi}{\sigma_d^2} \mathrm{tr} \left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf x_t \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace. \label{eq:Qfunctiontrace2}

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-(T-1)n_x\ln\sigma_d^2\nonumber \\
& -\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\right]\boldsymbol\theta

@@

&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  \mathbf q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)  \right] \boldsymbol\theta.

@@

The third term in \eqref{eq:Voldermont} is obtained using the same method in \cite{Shumway2000}, i.e. by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation of the second term in \eqref{eq:Qfunctiontrace1} . By defining $\Xi$-variables as 

@@

\boldsymbol\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\boldsymbol\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\label{eq:defofXi2}\\
\boldsymbol\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\right] \label{eq:defofXi3} \\	
\boldsymbol\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  \mathbf q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1} \mathbf q(\mathbf x_t)\right] \label{eq:defofXi4}\\
 \boldsymbol\Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)  \right] \label{eq:defofXi5}

@@

equation \eqref{eq:Voldermont} can be written in a form of

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_y\ln \sigma_{\epsilon}^2-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \boldsymbol\Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace 

@@

&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\boldsymbol\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\boldsymbol\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \boldsymbol\Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2}\boldsymbol \Xi_{5} \boldsymbol\theta.

@@

Covariance and cross-covariance terms in equation \eqref{eq:VoldermontwitXivariables}, consist of both linear terms, $\boldsymbol\Xi_0$, $\boldsymbol\Xi_1$ and $\boldsymbol\Xi_2$ and non-linear terms, $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and  $\boldsymbol\Xi_5$. Finding linear $\boldsymbol\Xi$-variables is straightforward as they can be computed directly using smoother outputs, i.e. $\hat{\mathbf x}_t^b $, $\mathbf P_t^b$ and $\mathbf M_t^b$ \cite{Gibsona2005}. To show how it can be done, the smoothed covariance matrix at time $t+1$ can be written as
 \mathbf P_{t+1}^b&=\mathbf E_{\Theta'}\left[(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)^{\top}\right]=\mathbf E_{\theta'}\left[\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]-\mathbf {\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\label{eq:EM-PrimeryXi0derivation1}\\
 \mathbf M_{t+1}^b&=\mathbf E_{\Theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t^b)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)^{\top}\right]= \mathbf E_{\Theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf {\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top},\label{eq:EM-PrimeryXi1derivation1}
consequently,
\begin{align}\label{eq:EM-Xi0derivation1}
 \mathbf E_{\Theta'}\left[\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\mathbf P_{t+1}^b+\mathbf {\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\\
 \mathbf E_{\Theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}^b+\mathbf {\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\label{eq:EM-Xi1derivation1},
substituting \eqref{eq:EM-Xi0derivation1} and \eqref{eq:EM-Xi1derivation1} in \eqref{eq:defofXi0} and \eqref{eq:defofXi1} respectively , gives the final formula for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$, i.e.

@@

\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:Xi1Gibson}.
The expression for $\boldsymbol\Xi_{2}$ can be derived in a similar manner, giving

@@

However calculation of non-linear $\boldsymbol\Xi$-variables is not straightforward as in linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $\mathbf q(\mathbf{x}_t)$ at each time instant which is to be used in  calculations of $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and $\boldsymbol\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^b$ at each time instant, i.e.

@@

 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^b\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right),

@@

\begin{align}\label{eq:SigmoidDerivative}
 f'(\boldsymbol \phi^\top\left(\mathbf r'\right)&\hat{\mathbf x}_t^b)= \frac{\varsigma}{\left(1 + \exp \left( \varsigma \left( v_0 - \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b \right) \right)\right)^2} \times \exp \left( \varsigma \left( v_0 - \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b \right) \right) \nonumber \\
=&\frac{\varsigma}{1 + \exp \left( \varsigma \left( v_0 - \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b \right) \right)} \times \left(1-\frac{1}{1 + \exp \left( \varsigma \left( v_0 - \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b \right) \right)}\right) \nonumber \\
=& \varsigma f(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b)\left( 1-f(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right).
substituting \eqref{eq:SigmoidTaylor} in \eqref{eq:QmatrixForSigmapoints},  an approximation for $\mathbf q(\mathbf x_t)$ around the smoothed state $\hat{\mathbf{x}}_t^b$ can be obtained
 \mathbf q(\mathbf x_t) \approx \mathbf q(\hat{\mathbf x}_t^b)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right) d\mathbf{r}'.

@@

Having approximated $\mathbf q\left(\mathbf x_t\right)$, by substituting  \eqref{eq:qTaylor} into equations (\ref{eq:defofXi3}-\ref{eq:defofXi5}) an approximation to the non-linear $\boldsymbol\Xi$-variables can be made. The three nonlinear variables are dealt with in turn. To start with, first consider equation \eqref{eq:defofXi3}, substituting in equation \eqref{eq:qTaylor} for $\mathbf q\left(\mathbf x_t\right)$, the term under the summation can be approximated by
\begin{align}\label{eq:Xi3Derivation1}
 \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t) &\approx \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)+\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\int_\Omega \boldsymbol{\Psi}(\mathbf{r}')\boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}',
noting that $\phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)$ is a scalar, equation \eqref{eq:Xi3Derivation1} can be rewritten as 
\begin{align}\label{eq:Xi3Derivation2}
 \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\approx\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)+\int_\Omega \boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}',
taking expected value of \eqref{eq:Xi3Derivation2} gives
\begin{align}\label{eq:Xi3Derivation3}
 \mathbf E_{\boldsymbol\Theta'}\left[\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\right] &\approx \mathbf E_{\boldsymbol\Theta'}\left[\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b) \nonumber \\
&+\int_\Omega \boldsymbol \phi^\top(\mathbf r')  \mathbf E_{\boldsymbol\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}',
Substituting for  $\mathbf E_{\boldsymbol\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_{t+1}^\top\right]$ from equation \eqref{eq:EM-PrimeryXi1derivation1} and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ yields
